---
title: "7.2 Data Visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)
```

---

## 7.2.1 Variable Types

---

### Bee Fertility Data

In this section we will be working with a data set from the following paper: 

Stuligross, C., & Williams, N. M. (2020). *Pesticide and resource stressors additively impair wild bee reproduction.* Proceedings. Biological sciences, 287(1935)

Stuligross & Williams look at the effect of pesticides and food resource stress on the fertility of bees. Bees lived in large flight cages with either a high or low density of wildflowers (in this data set this is referred to as "resources"), and the wildflowers were either treated or not treated with the pesticide imidacloprid. The number of offspring of the female bees was counted, along with the number of days they nested (i.e., days the bees actively produced offspring), and the intertegular span of female offspring (a measurement of the length between wings as a proxy for bee size). 

This dataset contains information for about 150 female bees. Columns are:
- `beeID`: a unique identifier for each female bee
- `cage`: the numeric identifier of the cage the bee lived in
- `pesticide`: whether wildflowers in cage were treated with pesticide ("treated" or "not treated")
- `resources`: density of wildflowers in cage ("high" or "low")
- `total.offspring`: number of offspring produced by bee
- `total.nesting.days`: number of days 
- `ITS`: intertegular span in centimetres

```{r}
bee_fertility <- read.csv("/home/jovyan/Week.7/7.2.Data.Visualization/bee_fertility_data.csv", stringsAsFactors = FALSE)
bee_fertility
```

---

### Variable Types

*Categorical variables*

Categorical variables have a finite number of values. A categorical value defines the category (group) that each item belongs to. In a data frame, columns containing categorical variables can be logical, character, or numeric type values. All logical type columns are categorical variables since there are only two distinct options. 

To determine if a character or numeric value is categorical, check if there is finite set of values that the variable can take. These values should be repeated multiple times.

We can use the `group_by()` and `summarise()` functions from the `dplyr` package to check: 

```{r}
library(dplyr)
bee_fertility %>% group_by(pesticide) %>% summarise(n=n())
```

We can see there are two possible values for `pesticide` and the number of rows that have each value.

Three columns in the `bee_fertility` data set are categorical variables:
- `cage`: the numeric identifier of the cage the bee lived in
- `pesticide`: whether wildflowers in the cage were treated with pesticide ("treated" or "not treated")
- `resources`: density of wildflowers in cage ("high" or "low")

Note that `cage` is a categorical variable because the numbers indicate which cage the bee is in; i.e., they indicate groups. The order of the numbers does not matter and the magnitude of the values has no meaning. To determine if numeric value is categorical, consider whether you could change the numbers to letters without losing any information.

*Discrete variables*

Discrete variables have a countable number of possible values (between a minimum and maximum value). In a data frame, columns containing discrete variables are always numeric type.

Two columns in the `bee_fertility` data set are discrete variables:
- `total.offspring`: number of offspring produced by bee
- `total.nesting.days`: number of days 

Both of these values must be whole natural numbers; a bee cannot produce 8.4282 offspring, thus it is a countable numeric value. Unlike the cage number, these numeric values are measurements and have meaning.

*Continuous variables*

Continuous variables have an infinite number of possible values (between a minimum and maximum value). In a data frame, columns containing continuous variables are always numeric type. 

Continuous variables differ from discrete variables in that values are not countable; they can take on any value in some range (i.e., numbers with decimals). Some general examples of continuous values are: expression level, heights and lengths, proportions.

Only one column in the `bee_fertility` data set is a continuous variable:
- `ITS`: intertegular span

*Identifiers*

Most datasets you will work with will also have a column with a identifier (ID). A unique identifier is a character value of numbers and/or letters that uniquely defines each sample. These are not classified as categorical, discrete, or continuous. 

There is one identifier in this data set:
- `beeID`: unique identifier for each female bee

---

## 7.2.2 Plot Types & 7.2.3 The ggplot2 Package

---

**Note that this section combines the information lecture sections 7.2.2 and 7.2.3**

---

### Introduction to ggplot2

Data visualization is an extremely important aspect of data analysis. While results of calculations and statistical tests are required understand the relationships between features and reach conclusions, plotting data helps us to visualize these relationships. Plots are also required for science communication via writing reports and papers and giving presentations.

The `ggplot2` package provides a set of functions for creating plots in R that give you more control over aspects of the graphical output than you can have using base R plotting functions.

Let's load the ggplot2 package, which has already been installed on the JupyterHub server:

```{r}
library(ggplot2)
```

The `ggplot2` package uses a different syntax from functions we have used before, but shares some similarities with `dyplr`. The general framework is:

    ggplot(data_frame, aes(column_name1, column_name2)) + geom_plot_type()
    
Within the first function `ggplot()` we provide the data frame that contains the data we want to plot. The second argument uses the function `aes()` (for "aesthetics"). The first `column_name1` provided to `aes()` has data for the x-axis, and the second `column_name2` provided to `aes()` has data for the y-axis. Finally, we use a `+` to add on a `ggplot2` function that defines the type of plot we want to create.

---

### Bar Plots

A *bar plot* (or bar chart) is used to visualize the distribution of samples across a set of categories (i.e., visualize a categorical variable). A bar is made for each category and the length of the bar visualizes the number (or count) of samples (observations) that belong to the category.

With respect to data frames, a bar plot shows the count of rows in a categorical data frame column that belongs to each of the categories in that column. 

Only one column is required in `aes()` for bar plots because the y-axis is a count which the `geom_bar()` function will determine automatically:

    ggplot(data_frame, aes(categorical_col)) + geom_bar()

For example, the `pesticide` column is a categorical variable: every element in the column is either "treated" or "not treated". We can make a bar plot for this column:

```{r}
ggplot(bee_fertility, aes(pesticide)) + geom_bar()
```

The plot shows the count of female bees that were in cages with wildflowers treated with pesticide and the number of bees in cages with wildflowers not treated with pesticide. 

---

### Histograms

A *histogram* is used to visualize the distribution of samples across a discrete or continuous variable. In a histogram, the minimum and maximum value for the variable are used to define a number of equally sized *bins*, and a bar is made for each bin showing how many values fall into that bin.

With respect to data frames, a histogram shows the count of rows in a discrete or continuous data column that falls within each of the bins defined for that column. 

Only one column is required in `aes()` for bar plots because the y-axis is a count which the `geom_histogram()` function will determine automatically:

    ggplot(data_frame, aes(discrete_or_continuous_col)) + geom_histogram(binwidth=NULL)

For example, the `total.offspring` column is a discrete variable. We can make a histogram with this column:

```{r}
ggplot(bee_fertility, aes(total.offspring)) + geom_histogram()
```

Notice that there was a warning at the top of our plot: "`stat_bin()` using `bins = 30`. Pick better value with `binwidth`."

It is warning us to pick a better *bin width*. The `geom_histogram()` function binned our data into 30 bins. To determine the current bin width for each of these 30 bins, we can calculate:

```{r}
(max(bee_fertility$total.offspring) - min(bee_fertility$total.offspring))/30
```

This means we have one bin for every 1.8 offspring in our histogram. This does not make much sense for our data since a bee cannot have 1.8 offspring. For discrete variables we should always set `binwidth` to 1 so that there is one bin for each possible value.

```{r}
ggplot(bee_fertility, aes(total.offspring)) + geom_histogram(binwidth = 1)
```

We can also make a histogram for the continuous value `ITS` which is a measurement of the length of the bee between the base of its wings. The values range from 0.263 to 0.359 centimetres. For a more detailed plot we can pick a bin width of 0.005:

```{r}
ggplot(bee_fertility, aes(ITS)) + geom_histogram(binwidth = 0.005)
```

Or, if we pick a larger bin width of 0.01, we can get a more general idea of the `ITS` distribution:

```{r}
ggplot(bee_fertility, aes(ITS)) + geom_histogram(binwidth = 0.01)
```

---

### Scatter Plots

*Scatter plots* are used to visualize the relationship between two numeric variables. The variables can be either discrete or continuous. In a scatter plot, a point is made on the plot for each sample (observation) showing the value of one variable along the x-axis and the value of the other variable along the y-axis.

Two data frame columns are required in `aes()` for scatter plots:

    ggplot(data_frame, aes(discrete_or_continuous_col1, discrete_or_continuous_col2)) + geom_point()
    
Let's examine the relationship between the number of offspring produced by the bees and the number of days spent nesting. Both are discrete variables.

```{r}
ggplot(bee_fertility, aes(total.offspring, total.nesting.days)) + geom_point()
``` 

We can see that these two variables, the number of offspring and the number of nesting days, look like they are correlated. When we are looking at the relationship between two continuous or discrete variables, the standard statistical test is the *Pearson correlation*, which can be calculated using the `cor()` function:

```{r}
cor(bee_fertility$total.offspring, bee_fertility$total.nesting.days)
```

---

### Box Plots

*Box plots* are used to visualize the relationship between a categorical variable and a continuous or discrete variable. There is one box for each category in the categorical variable along the x-axis, and the distribution of the continuous or discrete variable is shown via a box and whiskers along the y-axis based on a five point summary. 

Before we take a closer look at what a box represents, let's view an example box plot. 

Two columns are required in `aes()` for box plots: `geom_boxplot()`:

    ggplot(data_frame, aes(categorical_col, discrete_or_continuous_col)) + geom_boxplot()
    
Let's examine the distribution of the number of offspring produced by bees that were in cages not treated with pesticides vs. the number of offspring produced by bees in cages with pesticide treated flowers:

```{r}
ggplot(bee_fertility, aes(pesticide, total.offspring)) + geom_boxplot()
```

On the x-axis we have the two categories ("not treated" & "treated") and we can see the distribution of the number of offspring along the y-axis represented by boxes.

To understand what the boxes represent, first we should review the following definitions:
- *First Quartile*: the value in a set of observations which 25% of all values fall below.
- *Third Quartile*: the value in a set of observations which 75% of all values fall below.
- *Interquartile Range* (IQR): the magnitude of the difference between the first and third quartiles.

Box Features:
- The mid line of the box indicates the median value of the distribution. 
- The bottom and top of the box indicate the first and third quartiles respectively. 
- The lower whisker starts at the minimum value that is greater than or equal to:
    `first quartile - 1.5 x IQR`
- The upper whisker ends at the maximum value that is less than or equal to:
    `third quartile + 1.5 x IQR`
- Values outside of the whiskers are considered *outliers* and are represented by points.

Remember that we can get many of these values for a vector using the `summary()` function:

```{r}
summary(bee_fertility %>% filter(pesticide == "treated") %>% select(total.offspring))
```

If we look back at the plot above, it appears that the median number of offspring produced by female bees in the absence of pesticides is much higher than in the presence of pesticides.

To compare the distribution of a continuous or discrete variable between two categories, the standard statistical test is a *t-test*, which can be calculated with the `t.test()` function:

```{r}
t.test(bee_fertility %>% filter(pesticide == "treated") %>% select(total.offspring),
       bee_fertility %>% filter(pesticide == "not treated") %>% select(total.offspring))
```

The p-value is significant, leading to the conclusion that pesticides negatively affect the reproductive capability of female bees.

---

## 7.2.4 Adding Dimensions

---

### Visualizing Two Categorical Variables

To visualize counts for a single categorical variable we can use a bar plot. To visualize counts across two categorical variables we can use a stacked bar plot or a side-by-side bar plot. 

In the stacked bar plot, the height of the bar indicates how many samples are in the primary category, and colour indicates how many are in each of the categories of the secondary categorical variable. 

How do we do this ggplot? The `aes()` function has an optional argument called `fill` which takes the name of a categorical column and fills sections of the bar plot with different colours accordingly:

    ggplot(data_frame, aes(categorical_col1, fill=categorical_col2)) + geom_bar()

Let's remake a bar plot of the pesticide variable, but change the fill colour to the categorical variable "resources":

```{r}
ggplot(bee_fertility, aes(pesticide, fill=resources)) + geom_bar()
```

We can see that in place of the gray bars are two bars that are each split into two parts to indicate how much of each bar belongs to the secondary category ("resources").

The same information can be displayed side-by-side instead of stacked. Each of the categories for the primary categorical variable will have a bar for each of the categories in the secondary categorical variable.

The `geom_bar()` function has an argument called `position` with a default value of "stack". If it is changed to "dodge", it will create a side-by-side two-category bar plot: 

    ggplot(data.frame, aes(categorical_col1, fill=categorical_col2)) + geom_bar(position="dodge")

```{r}
ggplot(bee_fertility, aes(pesticide, fill=resources)) + geom_bar(position="dodge")
```

---

### Adding Categories to Scatter & Box Plots

We can also add an extra categorical variable to a box plot using the fill argument: 

    ggplot(data_frame, aes(categorical_col, continuous_or_discrete_col, fill=categorical_col)) + geom_boxplot()

Let's make a box plot that shows the distribution of total offspring across the pesticide and resource categories:

```{r}
ggplot(bee_fertility, aes(pesticide, total.offspring, fill=resources)) + geom_boxplot()
```

We can see that there are two boxes for each of the pesticide categories (primary categorical variable), one for each of the resource categories (secondary categorical variable). 

Viewing the distributions of the pesticide treated samples with high and low resources, we can see that the amount of resources had little affect on the number of offspring female bees produced. 

However, when we look at the samples not treated with pesticides, we see a large difference in the distributions between bees with high and low resources. The median number of offspring produced by female bees with low food resources is nearly 10 counts lower than the the median number of offspring produced by female bees with high food resources.

---

Scatter plots do not have a box or bar to fill using a categorical variable. However, we can change the colour of the individual points based on a categorical variable. This is done with the "colour" argument in the `aes()` function:
    
    ggplot(data_frame, aes(continuous_or_discrete_col, continuous_or_discrete_col, colour=categorical_col)) + geom_point()
    
Let's colour a scatter plot of total nesting days vs. total offspring using the pesticide categories:

```{r}
ggplot(bee_fertility, aes(total.offspring, total.nesting.days, color=pesticide)) + geom_point()
```

It is not always easy to see trends in this type of plot, however we can see that there are more blue "treated" dots in the bottom left hand corner and more pink/red "not treated" dots in the top right hand corner. This suggests that the pesticide-treated samples tended to have lower numbers of nesting days and offspring.

---

## 7.2.5 Plot Formatting

---

### Labels & Axes

It is good practice to have a plot title describing the data in the plot, and to have axis and legend labels that clearly describe the variables. 

The ggplot package has the `labs()` function for adding labels: 

    ggplot(...) + geom_plot_type(...) + labs(title=", x="", y="", fill="", colour="")

Titles are added with the `title` argument, and x and y axis labels with the `x` and `y` arguments. 

Legend labels can be added using the `fill` and `colour` arguments. These arguments will only work if you have used the `aes()` `fill` or `colour` arguments.

For example:

```{r}
ggplot(bee_fertility, aes(pesticide, fill=resources)) + 
    geom_bar(position="dodge") + 
    labs(title="Number of Female Bees Per Condition", 
         x="Pesticide Exposure", y="Number of Bees", 
         fill="Food Resources")
```

Looking at this plot with the more descriptive labels, it is easier to see how many bees were under each condition in the experiment, and to understand what the conditions were.

We can also change the formatting of plots by altering the axis limits. The axis limits are the minimum and maximum values of the axes. This is done with functions `xlim()` and `ylim()`, each of which take a vector of length two as an argument, detailing where the axis should start and end: 

    ggplot(...) + geom_plot_type(…) + xlim(c(start, end)) + ylim(c(start, end))

How can we use these functions to create a plot that shows the number of offspring for bees with fewer than 20 nesting days? If we put `total.nesting.days` on the y-axis and we only want to show the days from 0 to 20, we can alter the range of the axis with the `ylim()` function:

```{r}
ggplot(bee_fertility, aes(total.offspring, total.nesting.days)) + 
    geom_point() + 
    ylim(c(0,20)) + 
    labs(title="Number of Offspring for Bees with < 20 Nesting Days")
```

---

### Formatting Scatter Plots

The clarity of scatter plots can be improved by changing the appearance of the plot points. As points can often overlap, we can alter the opacity of the points using the `alpha` argument in `geom_point()`. The size of points can also be altered with the `size` argument: 

    ggplot(...) + geom_point(alpha=1, size=1.5)

`alpha` must be between 0 and 1 (0 is invisible, 1 is opaque/solid)
`size` has a lower limit of 0 

First let's recreate the plot showing the total nesting days vs. the total number of offspring for each bee coloured by the resources category to see how it looks without any other formatting:

```{r}
ggplot(bee_fertility, aes(total.offspring, total.nesting.days, colour=resources)) + 
    geom_point()
```

Now let's change the opacity and size of the points by setting `alpha` to 0.5 and `size` to 3:

```{r}
ggplot(bee_fertility, aes(total.offspring, total.nesting.days, colour=resources)) + 
    geom_point(alpha=0.5, size=3)
```

We can see that the points are much larger, but we can now also see that in the bottom left-hand corner there are many overlapping points at coordinate `0,0`, which would not have been evident if the points were fully opaque - we would only see the point on top.

There are many other functions and techniques for altering the formatting of a plot made with the `ggplot2` package. Check out the following links to learn more:

https://ggplot2.tidyverse.org/
https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf

---

### Saving Plots

Plots you create can be saved as a PDF, PNG, JPEG, or other graphics file (ex. TIFF). Each of these have their own function:
- Save a PDF: `pdf(filename, width=7, height=7)`
- Save a PNG: `png(filename, width=480, height=480)`
- Save a JPEG: `jpeg(filename, width=480, height=480)`

Each takes a file name and optional width and height arguments. PDF width and height values are in inches, whereas PNG and JPEG width and height values are in pixels. 

How are these functions used? First, the plot you want to write to a file must be saved to a variable. Then use the function for the type of plot you would like to create, print the variable containing the plot, and finally call the function `dev.off()` which closes the graphics device after the plot is written to the file:

    plot_variable <- ggplot(...) + ...
    pdf(...)
    print(plot_variable)
    dev.off()
    
For example, we can write a scatter plot to a PDF file:

```{r}
g <- ggplot(bee_fertility, aes(total.offspring, total.nesting.days)) + geom_point()
pdf("bee_fertility_offspring_vs_nesting_days.pdf", width=5, height=5)
print(g)
dev.off()
```

All we see as output is `null device 1`, which indicates the graphics device was closed properly. If you look in your current working directory you should now see a file called `bee_fertility_offspring_vs_nesting_days.pdf`. 

You can use relative or absolute paths in the `filename` argument to save your plot to different directories as needed.

---
