---
title: "7.1 Function Creation - Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
library(dplyr)
```
**Run the above block to load required packages**

---

At the end of this tutorial you should be able to:

* Write a function with the correct syntax
* Use required and optional arguments in a function
* Add warning messages to a function
* Add error messages to a function
* Determine how to implement a function for specific applications

---

*Note:* Even where not stated, you are encouraged to test your functions before checking them with otter grader.

---

## 7.1.1: Function Creation Basics

---

Recreate the `sum()` function. Call your function "my_sum_function". It should take a single argument `x` which is a vector of numbers and return the sum. Don't use the built-in `sum()` function.

```{r}
# Write and execute your code here

```

Before you check your code with otter grader, use the empty code chunk below to try your function. Here are some examples of what your function should return with different inputs:

    my_sum_function(1:10)
    [1] 55
    
    my_sum_function(c(109, 205, 1034))
    [1] 1348
    
```{r}
# Try your function here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.1.a.R")
```
**Check your answer by running the above block**

---

For the next portion of this tutorial we will use data from the Pan-cancer analysis of whole genomes (ICGC/TCGA, Nature 2020). This data set includes patient tumor samples for X patients across X types of cancer. For each patient/sample, the sex, age, survival status, general cancer type, detailed cancer type, and the "mutational burden" is included. Mutational burden is the number of non-synonymous mutations per megabase. There are also columns for 14 cancer associated genes indicating how many mutations the patient has in the gene. The genes are: TP53, KRAS, LRP1B, PCLO, PIK3CA, KMT2D, ARID1A, FAT4, CTNNB1, CDKN2A, APC, KMT2C, PTEN, BRAF.

```{r}
cancer_patient_data <- read.table("/home/jovyan/Week.7/7.1.Functions/Tutorial.7.1/cancer_patient_data.txt", stringsAsFactors = FALSE, header=TRUE, sep="\t")
cancer_patient_data
```

Write a function to identify the total number of patients with at least one mutation in a given gene. The function should be called "num.patients.w.mutation" and should take a single argument "gene" which should be the name of the gene. 

```{r}
# Write and execute your code here

```

Before you check your code with otter grader, use the empty code chunk below to test your function. Here are some examples of what your function should return with different inputs:

    num.patients.w.mutation("TP53")
    [1] 849

    num.patients.w.mutation("PCLO")
    [1] 208

```{r}
# Try your function here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.1.b.R")
```
**Check your answer by running the above block, but test it below first** 

---

## 7.1.2: Function Arguments

---

Write a function to identify the total number of patients with N mutations in a given gene. The function should be called "patients.w.N.mutations" and should take two arguments: "gene" which should be the name of the gene, and "N" which should be an integer (0,1,2,etc.). 

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.2.a.R")
```
**Check your answer by running the above block** 

Write a function identical to the above, except that "N" should be a optional argument with a default value of 1. Call the function "patients.with.N.mutations.2".

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.2.b.R")
```
**Check your answer by running the above block** 

---

Write a function that calculates the mean age of patients with a particular sex, survival status, and cancer type. Age should be rounded to 0 digits after the decimal place. Choose your own names for the function and arguments. Note that the dplyr package has been loaded for you. 
Hint: There are many `NA` values in this data frame, the `mean()` function has an argument called `na.rm` which can be set to `TRUE` to deal with `NA` values in the calculation.

(This code block will not be autograded).

```{r}
# Write and execute your code here

```

The following questions will use otter grader tests to ensure your function works.

Apply your function to get the mean age of female patients who are living and have breast cancer ("Breast Cancer"). Assign the function output to a variable called "mean_age_test_1".

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.2.c.R")
```
**Check your answer by running the above block** 

Apply your function to get the mean age of male patients who are deceased and had pancreatic cancer ("Pancreatic Cancer"). Assign the function output to a variable called "mean_age_test_2".

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.2.d.R")
```
**Check your answer by running the above block** 

---

## 7.1.3 Function Testing

---

Write a function called "is_seq" that checks if a provided character value (call the argument "seq") is a DNA sequence or an RNA sequence. The function should have an optional argument called "seq.type" with a default value of "DNA". If `seq.type="DNA"` then it should return `TRUE` if the provided character value is a DNA sequence, and `FALSE` otherwise. If `seq.type="RNA"`, it should instead check if the provided value is an RNA sequence and return `TRUE` or `FALSE` appropriately.

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.3.a.R")
```
**Check your answer by running the above block** 

Copy your code for the `is_seq()` function above into the code chunk below. Change the name of the function to "is_seq2". In this version of the function there will be warnings and errors. Add warnings and errors for the following (write your own warning and error messages):

- The function should return `FALSE` AND output a warning if `seq` is not a character value.
- The function should throw an error if `seq.type` is not equal to "DNA" or "RNA".

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.3.b.R")
```
**Check your answer by running the above block** 

Call your `is_seq2()` function using the character value `DNA.sequence` below. Set the `seq.type` argument so that it will return FALSE. Assign the function output to a variable called "is_seq2_FALSE":

```{r}
DNA.sequence <- "CGTATCGGCTATGCCTAGCACCG"
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.3.c.R")
```
**Check your answer by running the above block** 

Call your `is_seq2()` function using so that it will return FALSE and output a warning. Assign the function output to a variable called "is_seq2_FALSE_warning":

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.3.d.R")
```
**Check your answer by running the above block** 

Call your `is_seq2()` function using arguments that will cause it to throw an error. Assign the function output to a variable called "is_seq2_error":

```{r}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Tutorial.W7.1.3.e.R")
```
**Check your answer by running the above block** 

---